Metropolitan Statistical Area Evaluation

### 2022-11-18 src/demo_data_wrangling.Rmd
output code directly to google sheet that is used for the scoring Rmds

library(tidyverse)
library(here)
library(janitor) #clean_names
library(DT) # create data table
library(scales) 
library(lubridate)
library(gridExtra) 
library(ggrepel)
library(kableExtra)

library(tigris) # Census shape files
library(sf)
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
```

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2019) %>% 
  full_join(income_potential_MSAs) %>% 
  drop_na() %>% 
  select(cbsa, MSA, year, month, employment_rate)

@@

yoy_test <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  filter(year >= "2020") %>% 
  pivot_wider(names_from = year,
              values_from = avg_employment_rate,
              names_prefix = "avg_employment_rate") %>% 
  mutate(avg_emp_yoy = (avg_employment_rate2021 - avg_employment_rate2020)/avg_employment_rate2020 *100)

@@

```{r median rent}
# median rents from hud
MSA_median_rent2021_data <- read_csv(here("data", "raw", "2021MSA_medianrent.csv")) %>% 
  separate(col = cbsasub21,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = as.double(cbsa)) %>% 
  mutate(median_rent2021 = as.numeric(rent50_0)) %>% 
  select(cbsa, median_rent2021) %>% 
  right_join(potential_MSAs)

MSA_median_rent2022_data <- read_csv(here("data", "raw", "2022MSA_median_rent.csv")) %>% 
  separate(col = cbsasub22,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = as.double(cbsa)) %>% 
  mutate(median_rent2022 = as.numeric(rent_50_0)) %>% 
  select(cbsa, median_rent2022) %>% 
  right_join(MSA_median_rent2021_data)

@@

metro_house_prices <- metro_house_cost_data %>%
  select(cbsa, house_price_2021) %>% 
  right_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021, median_income, avg_emp_YoY, median_rent2021, median_rent2022, house_price_2021) %>% 
  filter(popestimate2021 >= 500000)
```

```{r house price exploratory, eval = FALSE}
ggplot(data = metro_house_cost_data, aes(x = log(x2020))) +
   geom_histogram(bins = 100)

ggplot(data = metro_house_cost_data, aes(sample = log(x2020))) +
  geom_qq()

@@

- Augusta-Richmond County
- Clarksville, TN-KY
- Greeley, CO
- Killeen, TX
- Lubbock, TX
- McAllen-Edinburg-Mission, TX
- Salisbury, MD
- Savannah, GA

@@

### Property Taxes
It was difficult to find property tax information by city that was consistent across cities. Median real estate taxes in dollars are provided by the U.S. Census and data is collected at the MSA level. A metric of property tax and median home prices can be used to normalized.

Dataset: B25103  MORTGAGE STATUS BY MEDIAN REAL ESTATE TAXES PAID (USD) in 2019

Source:[U.S. Census Bureau](https://data.census.gov/cedsci/table?q=property%20tax&g=0100000US%2431000M1&d=ACS%201-Year%20Estimates%20Detailed%20Tables&tid=ACSDT1Y2019.B25103&tp=true)

```{r property taxes data wrangling, eval = FALSE}
property_tax_data <- read_csv(here("data","raw", "wide_property_tax_data.csv"))

# create new data frame with tax estimate values
median_tax_df <- property_tax_data %>% 
  select(ends_with("estimate")) %>% 
  slice(2) %>% 
  stack() %>% # flip data table to be tidy
  rename(MSA = ind) %>% 
  mutate(median_tax_USD = as.numeric(values)) %>% 
  separate(col = MSA,
           into = c("MSA", "state"),
           sep = ", ") %>% 
  separate(col = state,
           into = c("state"),
           sep = " ") %>% 
  select(!values)

# make data frame with margin of error values
moe_tax_df <- property_tax_data %>% 
  select(ends_with("error")) %>% 
  slice(2) %>% 
  stack() %>% # flip data table to be tidy
  rename(MSA = ind) %>% 
  separate(col = values,
           into = c("value", "tax_moe_USD"),
           sep = 1) %>% 
  mutate(tax_moe_USD = as.numeric(tax_moe_USD)) %>% 
  separate(col = MSA,
           into = c("MSA", "state"),
           sep = ", ") %>% 
  separate(col = state,
           into = c("state", "area"),
           sep = " ") %>% 
  select(!c(value, area))

MSA_property_tax <- metro_house_prices %>% 
  full_join(median_tax_df) %>% 
  left_join(moe_tax_df) %>% 
  drop_na(cbsa) # drop all MSA rows if NA in cbsa column

```{r property tax visual, fig.height = 7, fig.align='center', eval = FALSE}
ggplot(MSA_property_tax, aes(x = median_tax_USD, y = reorder(MSA, median_tax_USD))) +
  geom_errorbarh(aes(xmin = median_tax_USD - tax_moe_USD, xmax = median_tax_USD + tax_moe_USD)) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median property tax in 2019", 
       subtitle = "", 
       x = "Median Property Tax estimate (in USD)", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())
```

@@

```{r data table}
# table seperated by MSAs
datatable(metro_house_prices, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)
```

@@

```{r MSAs seperate into cities}
#  split up MSAs into cities to collect utility, rent, and landlord data.
msa_seperated <- metro_house_prices %>% 
  separate_rows(MSA, sep = "-") %>% 
  rename(city = MSA)# %>% # seperate MSA cities into their own columns
 # full_join(payroll_data, by = c("MSA" = "area")) %>% 
 # drop_na(pop_percent_change20_21)

>>>>

library(tidyverse)
library(here)
library(janitor) #clean_names
library(DT) # create data table
library(scales) 
library(lubridate)
library(gridExtra) 
library(ggrepel)
library(kableExtra)
library(googlesheets4)
library(tigris) # Census shape files
library(sf)
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
```

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2019) %>% 
  mutate(cbsa = ifelse(cbsa == '76750', "38860", cbsa)) %>% # change south portland, ME to corresponding cbsa
  mutate(cbsa = as.numeric(cbsa)) %>% 
  full_join(income_potential_MSAs) %>% 
  drop_na() %>% 
  select(cbsa, MSA, year, month, employment_rate)

@@

@@

```{r median rent}
# median rents from hud
MSA_median_rent2021_data <- read_csv(here("data", "raw", "2021MSA_medianrent.csv")) %>% 
  separate(col = cbsasub21,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = ifelse(cbsa == 'M6400', "38860", cbsa)) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% 
  mutate(median_rent2021 = as.numeric(rent50_0)) %>% 
  select(cbsa, median_rent2021) %>% 
  right_join(potential_MSAs)

MSA_median_rent2022_data <- read_csv(here("data", "raw", "2022MSA_median_rent.csv")) %>% 
  separate(col = cbsasub22,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = ifelse(unknown == '38860M', "38860", cbsa)) %>% 
  mutate(median_rent2022 = as.numeric(rent_50_0)) %>% 
  select(cbsa, median_rent2022) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% 
  right_join(MSA_median_rent2021_data)

@@

metro_house_prices <- metro_house_cost_data %>%
  select(cbsa, house_price_2021) %>% 
  right_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021, median_income, avg_emp_YoY, median_rent2021, median_rent2022, house_price_2021) %>% 
  mutate(house_price_2021 = ifelse(cbsa == "12260", "284900", house_price_2021)) %>% 
  mutate(house_price_2021 = ifelse(cbsa == "32580", "259000", house_price_2021))

MSAs <-metro_house_prices %>% 
  filter(popestimate2021 >= 500000)
```

@@

- Augusta-Richmond County
- Clarksville, TN-KY
- Greeley, CO
- Killeen, TX
- Lubbock, TX
- McAllen-Edinburg-Mission, TX
- Salisbury, MD

@@

@@

```{r data table}
# table seperated by MSAs
datatable(MSAs, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)

gs4_create("updated_MSAs")
sheet_write(MSAs, ss = "https://docs.google.com/spreadsheets/d/1lmTpSDwVANxdAg5sW87Q8l_gtig7M0EjPPJ6J5_7dwg/edit#gid=0", sheet = "MSAs")
```

@@

```{r MSAs seperate into cities}
#  split up MSAs into cities to collect utility, rent, and landlord data.
msa_seperated <- metro_house_prices %>% 
  separate_rows(MSA, sep = "-") %>% 
  rename(city = MSA) %>% # seperate MSA cities into their own columns
 drop_na(city)

gs4_create("updated_MSAs")
sheet_write(msa_seperated, ss = "https://docs.google.com/spreadsheets/d/1lmTpSDwVANxdAg5sW87Q8l_gtig7M0EjPPJ6J5_7dwg/edit#gid=0", sheet = "cities")
